Accelerating adaptive ultrasound algorithms using graphics processing units \vspace{0.5cm}

### 2014-04-27 ThesisJP/Background/background.tex
Finished comments from CI and my own from easter.

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{img/free_lunsh.png}
\caption{Illustration of transitions between different eras in the computer industry. Currently there are three ongoing transitions; multi-core, heterogeneous-core and cloud-core. Image courtesy of Herb Sutter (herbsutter.com).}
\label{fig:jungle}
\end{figure}

@@

In Section \ref{sec:gpgpu} a special type of instructions was mentioned, namely SIMD instructions. These are instructions which are designed to execute a single instruction across multiple data element in one clock cycle. At the end of the 20th century, Intel introduced a SIMD instruction set for their Pentium III processor called the Streaming SIMD Extensions (\nom{SSE}{Streaming SIMD extensions (instruction set)}). The first SSE instruction set had registers of 128 bit which meant that four single-precision (32 bit) floating point values could be processed per clock cycle. Updates of the SSE instruction set (SSE2, SSE3, SSE4, \nom{AVX}{Advanced vector extensions (instruction set)}, AVX2, and AVX-512) have added new instructions and longer registers. The newly introduced extensions, AVX and AVX2, add registers of 256 bit and fused multiply accumulate (\nom{FMA}{Fused multiply-accumulate (instruction)}) instructions  which means that sixteen single-precision floating point operations can be performed per clock cycle per core. A new extension, AVX-512, extends the registers to 512 bit and is found in the Intel accelerator board  Xeon Phi and might show up in the next generation of CPUs from Intel. The latest CPU architecture from Intel, the Haswell architecture, has two processing ports per core supporting FMA-AVX2 instructions, and is therefore capable of performing 32 single-precision floating point operations per clock cycle per core (Fig.\,\ref{fig:cpu_gpu}). For an eight-core\footnote{An 8-core Haswell CPU is scheduled to launch during 2014. The numbers in Fig.\,\ref{fig:cpu_vs_gpu} for the last three generations of CPUs are all based on 8-core CPUs.} CPU with a clock frequency of 3.8 GHz this sums to 
\begin{equation}
32 \times 8 \times 3.8 = 972.8\,\text{GFLOPS}.
\end{equation}
Hence, a theoretical peak performance of close to one tera single precision floating point operations per second (TFLOPS). This is the same number as visualized in Fig\,\ref{fig:cpu_vs_gpu} for the Haswell architecture (Single precision). In addition, the most powerful integrated GPU found on Haswell CPUs, the HD 5200, has 40 execution units capable of processing sixteen single-precision floating point values at 1.3 GHz. This adds
\begin{equation}
16 \times 40 \times 1.3 = 832\,\text{GFLOPS}
\end{equation}
in extra processing power to the APU.

Modern GPUs can also be interpreted as units consisting of multiple SIMD cores (Fig.\,\ref{fig:cpu_gpu}). The recent Kepler architecture from Nvidia can provide up to 15 SIMD cores\footnote{Nvidia likes to refer to the unit capable of processing one single precision floating point value as a core. In Nvidia's world, Kepler therefore has $15\times192=2880$ cores.} capable of processing 192 single-precision floating point values per clock cycle per core. These GPUs are also capable of performing FMA instructions, effectively doubling the number of theoretical FLOPS. The maximum theoretical peak single precision FLOPS for this architecture with a core frequency of 745 MHz is then found to be
\begin{equation}
15 \times 192 \times 2 \times 745 = 4291\,\text{GFLOPS}.
\end{equation} 
Each SIMD core also includes 16 special function units (\nom{SFU}{Special function unit (sin, exp, sqrt, ect.)}) which add close to 1 TFLOPS in additional computational power. This adds up to a total of 5 TFLOPS as shown in Fig.\,\ref{fig:cpu_vs_gpu} for the GeForce GTX 780 TI GPU. From Fig.\,\ref{fig:cpu_gpu} we see how the high number of SIMD cores found on GPUs and their larger width, compared with the SIMD cores on CPUs, results in five times the computational performance of a high-end CPU (without the integrated GPU).

@@

It is important to understand that not all algorithms will benefit from wider SIMD registers. As the width increase, the only algorithms that will benefit are those which can be divided into a sufficiently large number of independent calculations. Also, to achieve peak performance, the implementation must perform one FMA operation on each core per clock cycle, and it must be possible for the GPU to hide all memory latency by scheduling threads which are not stalled by a memory fetch. All of this is something which is impossible to achieve in practical applications, and therefore one will always end up with a throughput which is a fraction of the peak throughput.

@@

To reduce the processing requirement of the Capon beamformer for medical ultrasound imaging, several simplified version have been proposed in the literature \cite{Asl2012, Kim}. Among the most interesting proposals we find the low complexity adaptive (LCA) beamformer by Synnev\aa{}g \textit{et al.} \cite{Synnevag2011} and the beamspace (BS) Capon beamformer by Nilsen and Hafizovic \cite{Nilsen2009}. In medical ultrasound imaging beamspace typically refers to the polar grid that cardiac ultrasound data are located in prior to scan conversion. In combination with the Capon beamformer, beamspace refers to the wavenumber-space representation of the recorded signals, i.e., the spatial Fourier transform of the element data: 
\begin{align}
\vec{x}_{\text{BS},l} &= \mat{B}[x_l[n],\,\dotso,\,x_{l+L}[n]]^T\\
[\mat{B}]_{p,q} &= \frac{1}{\sqrt{L}}e^{-j 2 \pi p q / L}.
\end{align}
The idea behind BS-Capon beamforming is that spatial beams with little information can be removed from $\mat{\hat{R}}$ by removing rows in $B$. When investigating the complexity of the BS-Capon beamformer [\textbf{Paper II}] we find that it is dominated by the transformation of data into beamspace, $O(N_LLN_b)$, where \nom{$N_b$}{Number of selected beams in beamspace (BS-Capon)} is the number of selected beams.

@@

The principle behind ray casting is depicted in Fig.\,\ref{fig:vr}. The projection of volumetric data is performed by casting a ray out from every pixel in the output image (Fig.\,\ref{fig:vr_rc}). Along each ray, samples are gathered from the volume and assigned color and opacity. For ultrasound visualization, the color intensity is dictated by the ultrasound backscatter intensity, and the actual color is often encoded with depth information in order to improve depth perception. An example of a volume rendering of cardiac ultrasound data is presented in \textbf{Paper IV}. The opacity, however, is governed by a user-controlled \textit{opacity transfer function} (\nom{OTF}{Opacity transfer function (volume rendering)}) as shown in Fig.\,\ref{fig:vr_otf}. For ultrasound visualization a monotonically increasing function is typically used, which means that below a certain threshold all samples will be rendered transparent (opacity = 0). We will refer to this as the \textit{opacity threshold}, and for the example in Fig.\,\ref{fig:vr_otf} this threshold is 25.

The iterative composition scheme for front-to-back ray casting of one ray, from $s_0$ to $s_{N_s-1}$ where $s$ is the sample position and $N_s$ is the number of samples along the ray, is \cite{rtvg2006}
\begin{align}
I(s_0, s_{i+1}) &= I(s_0, s_{i}) + (1 - \alpha_{\text{acc},i}) c_{i+1}\\
\alpha_{\text{acc},i+1} &= \alpha_{\text{acc},i} + (1 - \alpha_{\text{acc},i}) \alpha(c_i+1).
\label{eq:fronToBack}
\end{align}
Here, $\alpha(v)$ is the opacity transfer function, $c_{k}$ is the color intensity of sample $s_{k}$, $\alpha_{\text{acc}}$ is the accumulated opacity, and $I(s_0, s_{N_s-1})$ is the intensity value of all the projected samples combined. If $\alpha(v)$ is close to one for one or two consecutive samples, $\alpha_{\text{acc}}$ will quickly approach a value of one. An $\alpha_{acc,k}$-value close to one means that the contribution from $c_{k+1}, \cdots, c_{N_s-1}$ to the final value, $I(s_0, s_{N_s-1})$, will be minimal. Hence, the iteration can stop at $i=k$ without loosing visible samples. This optimization technique is known as \textit{early-ray-termination}, and will significantly speed up the rendering algorithm if an opaque material is rendered.

@@

\section{Concluding remarks}
All the presented problems in this chapter, except DAS beamforming, have in common that they require a processing throughput in order of GFLOPS. Comparing this to the typical clock frequency of modern CPUs at 3 GHz we see that GFLOPS problems will require parallel execution in order to achieve real-time performance. With CPU threads the theoretical peak performance on an eight-core CPU increases to 24 GFLOPS. However, to reach a peak performance above 100 GFLOPS, SIMD instructions on a CPU or SIMT instructions on a GPU have to be used. It also must be possible to divide the problem into a large number of independent calculations.

>>>>

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{img/free_lunsh.png}
\caption{Illustration of transitions between different eras in the computer industry. Currently there are three ongoing transitions; multi-core, heterogeneous-core and cloud-core (herbsutter.com).}
\label{fig:jungle}
\end{figure}

@@

In Section \ref{sec:gpgpu} a special type of instructions was mentioned, namely SIMD instructions. These are instructions which are designed to execute a single instruction across multiple data element in one clock cycle. At the end of the 20th century, Intel introduced a SIMD instruction set for their Pentium III processor called the Streaming SIMD Extensions (\nom{SSE}{Streaming SIMD extensions (instruction set)}). The first SSE instruction set had registers of 128 bit which meant that four single-precision (32 bit) floating point values could be processed per clock cycle. Updates of the SSE instruction set (SSE2, SSE3, SSE4, \nom{AVX}{Advanced vector extensions (instruction set)}, AVX2, and AVX-512) have added new instructions and longer registers. The newly introduced extensions, AVX and AVX2, add registers of 256 bit and fused multiply accumulate (\nom{FMA}{Fused multiply-accumulate (instruction)}) instructions  which means that sixteen single-precision floating point operations can be performed per clock cycle per core. A new extension, AVX-512, extends the registers to 512 bit and is found in the Intel accelerator board  Xeon Phi and might show up in the next generation of CPUs from Intel. The latest CPU architecture from Intel, the Haswell architecture, has two processing ports per core supporting FMA-AVX2 instructions, and is therefore capable of performing 32 single-precision floating point operations per clock cycle per core (Fig.\,\ref{fig:cpu_gpu}). For an eight-core\footnote{An 8-core Haswell CPU is scheduled to launch during 2014. The numbers in Fig.\,\ref{fig:cpu_vs_gpu} for the last three generations of CPUs are all based on 8-core CPUs.} CPU with a clock frequency of 3.8 GHz this sums to 
\begin{equation}
32 \times 8 \times 3.8 = 972.8\,\text{GFLOPS}.
\end{equation}
Hence, a theoretical peak performance of close to one tera single precision floating point operations per second (TFLOPS). This is the same number as visualized in Fig.\,\ref{fig:cpu_vs_gpu} for the Haswell architecture (Single precision). In addition, the most powerful integrated GPU found on Haswell CPUs, the HD 5200, has 40 execution units capable of processing sixteen single-precision floating point values at 1.3 GHz. This adds
\begin{equation}
16 \times 40 \times 1.3 = 832\,\text{GFLOPS}
\end{equation}
in extra processing power to the APU.

Modern GPUs can also be interpreted as units consisting of multiple SIMD cores (Fig.\,\ref{fig:cpu_gpu}). The recent Kepler architecture from Nvidia can provide up to 15 SIMD cores\footnote{Nvidia likes to refer to the unit capable of processing one single precision floating point value as a core. In Nvidia's world, Kepler therefore has $15\times192=2880$ cores. A SIMD core on a Nvidia GPU is called a Streaming Multiprocessor (\nom{SM}{Streaming multiprocessor (SIMD core on Nvidia GPUs)}).} capable of processing 192 single-precision floating point values per clock cycle per core. These GPUs are also capable of performing FMA instructions, effectively doubling the number of theoretical FLOPS. The maximum theoretical peak single precision FLOPS for this architecture with a core frequency of 745 MHz is then found to be
\begin{equation}
15 \times 192 \times 2 \times 745 = 4291\,\text{GFLOPS}.
\end{equation} 
Each SIMD core also includes 16 special function units (\nom{SFU}{Special function unit (sin, exp, sqrt, ect.)}) which add close to 1 TFLOPS in additional computational power. This adds up to a total of 5 TFLOPS as shown in Fig.\,\ref{fig:cpu_vs_gpu} for the GeForce GTX 780 TI GPU. From Fig.\,\ref{fig:cpu_gpu} we see how the high number of SIMD cores found on GPUs and their larger width, compared with the SIMD cores on CPUs, results in five times the computational performance of a high-end CPU (without the integrated GPU).

@@

It is important to understand that not all algorithms will benefit from wider SIMD registers. As the width increase, the only algorithms that will benefit are those which can be divided into a sufficiently large number of independent calculations. Also, to achieve peak performance, the implementation must perform one FMA operation on each core per clock cycle, and it must be possible for the GPU to hide all memory latency by scheduling threads which are not stalled by a memory fetch. All of this is something that is not likely to happen in practical applications, and therefore one will always end up with a throughput which is a fraction of the peak throughput.

@@

To reduce the processing requirement of the Capon beamformer for medical ultrasound imaging, several simplified version have been proposed in the literature \cite{Asl2012, Kim}. Among the most interesting proposals we find the low complexity adaptive (LCA) beamformer by Synnev\aa{}g \textit{et al.} \cite{Synnevag2011} and the beamspace (BS) Capon beamformer by Nilsen and Hafizovic \cite{Nilsen2009}. In medical ultrasound imaging beamspace typically refers to the polar grid that cardiac ultrasound data are located in before it is scan converted into a Cartesian grid. In combination with the Capon beamformer, beamspace refers to the wavenumber-space representation of the recorded signals, i.e., the spatial Fourier transform of the element data: 
\begin{align}
\vec{x}_{\text{BS},l} &= \mat{B}[x_l[n],\,\dotso,\,x_{l+L}[n]]^T\\
[\mat{B}]_{p,q} &= \frac{1}{\sqrt{L}}e^{-j 2 \pi p q / L}.
\end{align}
The idea behind BS-Capon beamforming is that spatial beams with little information can be removed from $\mat{\hat{R}}$ by removing rows in $B$. When investigating the complexity of the BS-Capon beamformer [\textbf{Paper II}] we find that it is dominated by the transformation of data into beamspace, $O(N_LLN_b)$, where \nom{$N_b$}{Number of selected beams in beamspace (BS-Capon)} is the number of selected beams.

@@

The principle behind ray casting is depicted in Fig.\,\ref{fig:vr}. The projection of volumetric data is performed by casting a ray out from every pixel in the output image (Fig.\,\ref{fig:vr_rc}). Along each ray, samples are gathered from the volume and assigned a color and opacity. For ultrasound visualization, the color intensity is dictated by the ultrasound backscatter intensity, and the colorfulness is often encoded with depth information in order to improve depth perception. An example of a volume rendering of cardiac ultrasound data is presented in \textbf{Paper IV}. The opacity, however, is governed by a user-controlled \textit{opacity transfer function} (\nom{OTF}{Opacity transfer function (volume rendering)}) as shown in Fig.\,\ref{fig:vr_otf}. For ultrasound visualization a monotonically increasing function is typically used, which means that below a certain threshold all samples will be rendered transparent (opacity = 0). We will refer to this as the \textit{opacity threshold}, and for the example in Fig.\,\ref{fig:vr_otf} this threshold is 25.

The iterative composition scheme for front-to-back ray casting of one ray, from $s_0$ to $s_{N_s-1}$ where $s$ is the sample position and $N_s$ is the number of samples along the ray, is \cite{rtvg2006}
\begin{align}
I(s_0, s_{i+1}) &= I(s_0, s_{i}) + (1 - \alpha_{\text{acc},i}) c_{i+1}\\
\alpha_{\text{acc},i+1} &= \alpha_{\text{acc},i} + (1 - \alpha_{\text{acc},i}) \alpha(c_i+1).
\label{eq:fronToBack}
\end{align}
Here, $\alpha(v)$ is the opacity transfer function, $c_{k}$ is the color of sample $s_{k}$, $\alpha_{\text{acc}}$ is the accumulated opacity, and $I(s_0, s_{N_s-1})$ is the intensity value of all the projected samples combined. If $\alpha(v)$ is close to one for one or two consecutive samples, $\alpha_{\text{acc}}$ will quickly approach a value of one. An $\alpha_{acc,k}$-value close to one means that the contribution from $c_{k+1}, \cdots, c_{N_s-1}$ to the final value, $I(s_0, s_{N_s-1})$, will be minimal. Hence, the iteration can stop at $i=k$ without loosing visible samples. This optimization technique is known as \textit{early-ray-termination}, and will significantly speed up the rendering algorithm if an opaque material is rendered.

@@

\section{Concluding remarks}
All the presented problems in this chapter, except DAS beamforming, have in common that they require a processing throughput of several GFLOPS. Comparing this to the typical clock frequency of modern CPUs at 3 GHz we see that GFLOPS problems quickly require parallel execution in order to achieve real-time performance. With CPU threads the theoretical peak performance on an eight-core CPU increases to 24 GFLOPS. However, to reach a peak performance above 100 GFLOPS, SIMD instructions on a CPU or SIMT instructions on a GPU have to be used. However, to apply threads and SIMD instructions, it must be possible to divide the problem into a large number of independent calculations.
